---
---
#title: Merging and Cleaning all the three datasets
#author: "Team 02 (Aditya Dube and Richard Budden)"


#-#-# Merging and Cleaning all the three datasets #-#-#

      ## Step 0: Basic Setup - Install Packages / Load Libraries
      
      ## Step 1: Collecting / Explore Data
                 part a) Import Datasets
                 part b ) Clean Datasets
                 

      ## Step 2: Prepare Data
                 part a) Joining the Datasets
                 part b) Remove, Code, and/or Impute Data
                 part c) Variable Selection
          
                 
  
                  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


********************************************************************************
## Step 0: Basic Setup - Install Packages / Load Libraries
********************************************************************************
```{r Install_Load, message=F, error=F, eval=T, warning=F}
library(tidyverse)
library(tidymodels)
library("dplyr")
library(stringr)
```

********************************************************************************
## Step 1: Collecting / Explore Data     part a) Import Datasets
********************************************************************************

```{r Import_Datasets, message=F, error=F, eval=T, warning=F}
medicare_data <- read.csv("Medicare_Provider_Charge_Inpatient_2018_data.csv")
hospital_data <- read.csv("Hospital_General_Information.csv")
census_data_original <- read.csv("Census_Data_Original.csv")
census_data <- census_data_original #Hand off file so we can refer back to original

```


#Explore-Inspect the Dataset
```{r Inspect_Medicare_and_Hospital_Datasets, message=F, error=F, eval=T, warning=F}
head(medicare_data)
head(hospital_data)
#head(census_data_original)

```

********************************************************************************
## Step 1: Collecting / Explore Data     part b ) Clean Datasets
********************************************************************************

# Clean the zipcode entries and write file to folder
```{r Census_Data_Modify, message=F, error=F, eval=T, warning=F}
# Removing the first 5 characters in column 2 from row 2 onwards
census_data[2:nrow(census_data), 2] <- substring(census_data[2:nrow(census_data), 2], 7)

# Exclude the first row
census_data <- census_data[-1,]
write.csv(census_data, "Census_data_modified.csv", row.names = FALSE)
census_data_modified <- read.csv("Census_data_modified.csv")

```

#Explore-Inspect the Dataset
```{r Inspect_Census_Data_Modified, message=F, error=F, eval=T, warning=F}
#head(census_data_modified)

```

********************************************************************************
 ## Step 2: Prepare Data          part b) Joining the Datasets
********************************************************************************

```{r Merging_MIHPS_HGI, message=F, error=F, eval=T, warning=F}

MIHPS_HGI_Merged <- merge(medicare_data, hospital_data, by.x = "Rndrng_Prvdr_CCN", by.y = "Provider.ID")
write.csv(MIHPS_HGI_Merged, "MIHPS_HGI_Merged.csv", row.names = FALSE)

```

#Merge the Datasets
```{r Merging_Census_Data_MIHPS_HGI_Merge, message=F, error=F, eval=T, warning=F}
Final_data_original<- merge(MIHPS_HGI_Merged, census_data_modified, by.x = "Rndrng_Prvdr_Zip5", by.y = "NAME")
write.csv(Final_data_original, "Final_data_original.csv", row.names = FALSE)

```

#Explore-Inspect the Dataset
```{r Inspect_Final_Data_Original, message=F, error=F, eval=T, warning=F}
#head(Final_data_original)

```

## Cleaning The Final Data After Merging 
```{r Inspect_Data, message=F, error=F, eval=T, warning=F}
# Inspect data
#head(Final_data_original)
#tail(Final_data_original)
#summary(Final_data_original)

```


Clean the Datasets
```{r Clean_Data, message=F, error=F, eval=T, warning=F}

# remove rows with missing values
Final_data_cleaned <- Final_data_original[complete.cases(Final_data_original),]
# Handle duplicates
Final_data_cleaned <- unique(Final_data_original)
# Removing columns from 47 onwards
Final_data_cleaned <- select(Final_data_original, -47:-ncol(Final_data_original))
#  Rename the column
Final_data_cleaned <- rename(Final_data_cleaned, Pop_total = B16010_001E, Margin_error_poptotal = B16010_001M)
Final_data_cleaned <- rename(Final_data_cleaned, 
                                    Hospital_name = Hospital.Name, Zip_code = ZIP.Code, 
                                    County_name = County.Name, Phone_number = Phone.Number,
                                    Hospital_type = Hospital.Type, Hospital_ownership = Hospital.Ownership, 
                                    Emergency_services = Emergency.Services)

```


********************************************************************************
## Step 2: Prepare Data         part b) Remove, Code, and/or Impute Data
********************************************************************************

```{r Create_new_variable, message=F, error=F, eval=T, warning=F}
# Creating a new computed variable
Final_data_cleaned$cost_to_pat <- (Final_data_cleaned$Avg_Tot_Pymt_Amt - Final_data_cleaned$Avg_Mdcr_Pymt_Amt )

```


```{r Convert_Data_to_Numeric, message=F, error=F, eval=T, warning=F}
# convert variables to numeric type
Final_data_cleaned$Pop_total <- as.numeric(Final_data_cleaned$Pop_total)
Final_data_cleaned$Margin_error_poptotal <- as.numeric(Final_data_cleaned$Margin_error_poptotal)
# Creating a new computed variable
Final_data_cleaned$adjusted_pop <- (Final_data_cleaned$Pop_total - Final_data_cleaned$Margin_error_poptotal)

```


```{r save_to_file, message=F, error=F, eval=T, warning=F}
# saving it new csv file
write.csv(Final_data_cleaned, "Final_data_cleaned.csv", row.names = FALSE)

```

```{r Inpsect_Final_Data_Cleaned, message=F, error=F, eval=T, warning=F}
#head(Final_data_cleaned) 

```

********************************************************************************
## Step 2: Explore / Prepare Data     part c) Variable Selection
********************************************************************************

```{r AMI_Filtering, message=F, error=F, eval=T, warning=F}
Final_data_cleaned_AMI <- Final_data_cleaned %>% 
  filter(DRG_Cd %in% c("246", "247", "248", "249", "250", "251","280","281","282"))

#head(Final_data_cleaned_AMI)

```


```{r CABG_Filtering, message=F, error=F, eval=T, warning=F}
Final_data_cleaned_CABG <- Final_data_cleaned %>% 
  filter(DRG_Cd %in% c("231", "232", "233", "234", "235", "236"))
#head(Final_data_cleaned_CABG)

```

