---
title: Estimating Medicare Costs for all Codes
author: "Team 02 (Aditya Dube and Richard Budden)"
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
date: "3/20/2023"
output:
  pdf_document: default
  html documnet: default
---

#-#-# Predicting Medicare for All Codes #-#-#

      ## Step 0: Basic Setup - Install Packages / Load Libraries
      
      ## Step 1: Collecting Data
                 part a) Import Datasets
                 part b) Joining the Datasets

      ## Step 2: Explore / Prepare Data
                 part a) Remove, Code, and/or Impute Data
                 part b) Variable Selection
                 part c) Group by State/Summarize Data
                 
      ## Step 3: Visualization of Data
                 part a) Histogram
                 part b) United States Heat Map
                 
      ## Step 4: Creating Training and Test Datasets
      
      ## Step 5: Build and Evaluate Linear Regression Model

      ## Step 6: Build and Evaluate CART Model
      
      ## Step 7: Build and Evaluate Artifical Neural Network  Model (Feedforward ANN)
                  

********************************************************************************
                ## Step 0: Basic Setup - Install Packages / Load Libraries
********************************************************************************

```{r message=F, error=F, eval=T, warning=F}
library(dplyr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(maps)
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(neuralnet)

```

********************************************************************************
          ## Step 1: Collecting Data       part a) Import Datasets
********************************************************************************

```{r message=F, error=F, eval=T, warning=F}
# Load the three datasets
hospital_data <- read.csv("Hospital General Information.csv")
medicare_data <- read.csv("Medicare_Inpatient_Hospital_by_Provider_and_Service_2018_data.csv")
census_data <- read.csv("Census data.csv")
```

Initial look at the Structure of the Datasets

```{r message=F, error=F, eval=T, warning=F}
##  Structure of the dataset
str(hospital_data)
str(medicare_data)
str(census_data)

```

********************************************************************************
                ## Step 1: Collecting Data       part b) Joining the Datasets
********************************************************************************

Used the dplyr package in R to join multiple datasets based on a common variable.
We Joined the medicare and hospital data by Provider ID and by Zip code

```{r message=F, error=F, eval=T, warning=F}
final_data <- inner_join(medicare_data, hospital_data, by = c("Provider_ID","ZIP_Code"))

```


********************************************************************************
    ## Step 2: Explore / Prepare Data     part a) Remove, Code, and/or Impute Data
********************************************************************************
                 
Data cleaning: 
Before analyzing the data, we need to clean it by removing missing values, 
fixing formatting issues, and dealing with outliers

```{r message=F, error=F, eval=T, warning=F}
# Remove rows with missing values
final_data <- final_data[complete.cases(final_data), ]

```

********************************************************************************
          ## Step 2: Explore / Prepare Data     part b) Variable Selection
********************************************************************************

Data wrangling and variable selection: 
We use the tidyverse package in R to manipulate and select variables from our dataset.


```{r message=F, error=F, eval=T, warning=F}
# Select relevant variables

selected_data <- final_data %>%
  select("Provider_ID", "ZIP_Code","Ave_medical_payment", "Total_discharges", "Ave_covered_charges", "Ave_total_payment","Hospital.Type","Hospital.Ownership",
         "Hospital.overall.rating","State", "DRG_Cd")
attach (selected_data)

# Concerting character variable to string variables
selected_data[sapply(selected_data, is.character)] <- lapply(selected_data[sapply(selected_data, is.character)], 
                                       as.factor)

```


********************************************************************************
        ## Step 2: Explore / Prepare Data     part c) Group by State/Summarize Data 
********************************************************************************

```{r message=F, error=F, eval=T, warning=F}
# Calculate the average costs per state
Avg_Mdcr_Pymt_Amt  <- selected_data %>%
  group_by(State) %>%
  summarise(Avg_Mdcr_Pymt  = Ave_medical_payment)
Avg_Mdcr_Pymt_Amt <- Avg_Mdcr_Pymt_Amt[1:15537, ]

```


********************************************************************************
          ## Step 3: Visualization of Data     part a) Histogram 
********************************************************************************


```{r message=F, error=F, eval=T, warning=F}
# Create a histogram of average costs per state
# Create a histogram of average costs per state
ggplot(selected_data , aes(x = Ave_medical_payment)) +
  geom_histogram(binwidth = 10000) +
  labs(title = "Histogram of Average Medicare Payments", x = "Average Total Payments")
```


********************************************************************************
        ## Step 3: Visualization of Data     part b) United States Heat Map 
********************************************************************************

```{r message=F, error=F, eval=T, warning=F}
# Create a US density map of average costs per state
us_map <- map_data("state")
usmap <- cbind(us_map,Avg_Mdcr_Pymt_Amt)

head(usmap)

ggplot(usmap, aes(x = long, y = lat, group = group, fill = Avg_Mdcr_Pymt)) +
  geom_polygon(color = "black") +
  scale_fill_gradient(low = "white", high = "red", na.value = "lightgrey") +
  theme_void() + coord_map() +
  labs(title = "Average Medicare Payments by State (All Codes)")


```


********************************************************************************
          ## Step 4: Creating Training and Test Datasets
********************************************************************************


```{r message=F, error=F, eval=T, warning=F}
# Split the data into training and testing sets
set.seed(123)
library(caret)
trainIndex <- createDataPartition(selected_data$Ave_medical_payment, p = 0.7, list = FALSE)
train <- selected_data[trainIndex, ]
test <- selected_data[-trainIndex, ]
```


********************************************************************************
          ## Step 5: Build and Evaluate Linear Regression Model
********************************************************************************

```{r message=F, error=F, eval=T, warning=F}
# Build the linear regression model
model <- lm(Ave_medical_payment ~ Total_discharges + Ave_covered_charges + Ave_total_payment+ Hospital.overall.rating+Hospital.Ownership, data = train)

# Predict on the testing set
predictions <- predict(model, newdata = test)


# Check for the model assumptions
par(mfrow = c(2, 2))
plot(model)
```

Evaluate the performance of the Linear Model

```{r message=F, error=F, eval=T, warning=F}
RMSE(predictions, test$Ave_medical_payment)
R2(predictions, test$Ave_medical_payment)
```

********************************************************************************
          ## Step 6 Build and Evaluate CART Model (Regression Tree)
********************************************************************************

```{r}
# Build the regression tree model
model2 <- rpart(Ave_medical_payment ~ Total_discharges + Ave_covered_charges + Ave_total_payment+ Hospital.overall.rating+Hospital.Ownership, data = train, method = "anova")

#Plot
#install.packages("RGtk2")
# Plot the tree						
fancyRpartPlot(model2)

# Predict on the testing set
predictions <- predict(model2, newdata = test)

#Model 2
RMSE(predictions, test$Ave_medical_payment)
R2(predictions, test$Ave_medical_payment)



```


```{r}
printcp(model2)
plotcp(model2)

pfit = prune(model2, cp=model2$ptable[which.min(model2$cptable[,"xerror"]), "CP"])
pfit
```

```{r message=F, error=F, eval=T, warning=F}
# Build the regression tree model
control_setting <- rpart.control(minsplit = 2, cp = .005, xval = 10)

model2 <- rpart(Ave_medical_payment ~ Total_discharges + Ave_covered_charges + Ave_total_payment+ Hospital.overall.rating+Hospital.Ownership, data = train, method = "anova", control = control_setting)

# Plot
#install.packages("RGtk2")
fancyRpartPlot(model2)
# Predict on the testing set
predictions <- predict(model2, newdata = test)

RMSE(predictions, test$Ave_medical_payment)
R2(predictions, test$Ave_medical_payment)

```

Evaluate the performance of the CART model

```{r message=F, error=F, eval=T, warning=F}
#Model 2
RMSE(predictions, test$Ave_medical_payment)
R2(predictions, test$Ave_medical_payment)

```

Data pruning for Regression Tree

```{r message=F, error=F, eval=T, warning=F}

# Prune the tree, display pruned tree
library(tree)

tree.model <- tree(Ave_medical_payment ~ Total_discharges + Ave_covered_charges + Ave_total_payment+ Hospital.overall.rating+Hospital.Ownership, data = train)
prune.data <- prune.tree(tree.model)
plot(prune.data, col = "blue")

```



********************************************************************************
    ## Step 7 : Build and Evaluate Artifical Neural Network  Model (Feedforward ANN)
********************************************************************************
Feedforward ANN


```{r message=F, error=F, eval=T, warning=F}

## Scaling
scale_data <- final_data%>%
  select("Ave_medical_payment", "Total_discharges", "Ave_covered_charges", "Ave_total_payment")

set.seed(123)
trainIndex2 <- createDataPartition(scale_data$Ave_medical_payment, p = 0.7, list = FALSE)
train2 <- scale_data[trainIndex2, ]
test2 <- scale_data[-trainIndex2, ]

```


Min-Max Normalization and Scaling the input variable

```{r}
#transform your factor to numeric.
#tranform it to a factor and the to numeric
selected_data$Hospital.Ownership <- as.numeric(as.factor(Hospital.Ownership))
selected_data$Hospital.overall.rating <- as.numeric(as.factor(Hospital.overall.rating))

```

```{r message=F, error=F, eval=T, warning=F}

selected_data$Ave_medical_payment <- (selected_data$Ave_medical_payment - min(selected_data$Ave_medical_payment)) / (max(selected_data$Ave_medical_payment) - min(selected_data$Ave_medical_payment))

selected_data$Total_discharges <- (selected_data$Total_discharges - min(selected_data$Total_discharges)) / (max(selected_data$Total_discharges) - min(selected_data$Total_discharges))

selected_data$Ave_covered_charges <- (selected_data$Ave_covered_charges - min(selected_data$Ave_covered_charges)) / (max(selected_data$Ave_covered_charges) - min(selected_data$Ave_covered_charges))

 selected_data$Ave_total_payment <- (selected_data$Ave_total_payment - min(selected_data$Ave_total_payment)) / (max(selected_data$Ave_total_payment) - min(selected_data$Ave_total_payment))
 
selected_data$Hospital.overall.rating <- (selected_data$Hospital.overall.rating - min(selected_data$Hospital.overall.rating)) / (max(selected_data$Hospital.overall.rating) - min(selected_data$Hospital.overall.rating))
 
selected_data$Hospital.Ownership <- (selected_data$Hospital.Ownership - min(selected_data$Hospital.Ownership)) / (max(selected_data$Hospital.Ownership) - min(selected_data$Hospital.Ownership))
 

set.seed(123)
inp <- sample(2, nrow(selected_data), replace = TRUE, prob = c(0.7, 0.3))
training_data <- selected_data[inp==1, ]
test_data <- selected_data[inp==2, ]

```


```{r message=F, error=F, eval=T, warning=F}

#from RBloggers "Selecting the number of neurons in the hidden layer of a neural network"
#A Variation of this rule suggests to choose a number of hidden neurons between one and the number of Inputs minus the number of outputs

#Upon our Model with 5 Inputs and 1 Output we set Hidden Levels at 4 (5-1=4) 


set.seed(333)
model3 <- neuralnet(Ave_medical_payment ~ Total_discharges + Ave_covered_charges + Ave_total_payment+Hospital.Ownership+ Hospital.overall.rating,
               data = training_data,
               hidden = 4,
               linear.output = FALSE)

# Predict on the testing set
predictions <- predict(model3, test_data)

# Evaluate the performance of the model
RMSE(predictions, test_data$Ave_medical_payment)
R2(predictions, test_data$Ave_medical_payment)

# plot neural network
plot(model3, rep = "best")

```
